Deep learning for quality control of surface physiographic fields using satellite Earth observations

### 2023-04-30 docs/submission/template.tex
clean version, noise numbers required

\title{\hl{Deep learning for quality control of surface physiographic fields using satellite Earth observations}}

@@

%\introduction  %% \introduction[modified heading if necessary]
\section{Introduction}
\noindent Globally, there are $\sim$ 117 million lakes - defined as inland water bodies without lateral movement of water - making up around 3.7$\%$ of the Earth's land surface \citep{Verpoorter2014}. Their distribution is highly anisotropic, with the majority of lakes located between $45-75^{\circ}$N in the Boreal and Arctic regions. Lakes are highly important from the perspective of both numerical weather prediction and climate modelling \hl{as part of the EC-Earth model.}  For the latter, lakes generally influence the global carbon cycle as both sinks and sources of greenhouse gases; the majority of lakes are net heterotrophic, over saturated with CO$_2$ as a result of in lake respiration and so emit carbon into the atmosphere \citep{Pace2005,Tranvik2009}.  Total CO$_2$ emission from lakes is estimated at $1.25 - 2.30$ Pg of CO2-equivalents annually \citep{DelSontro2018}, nearly $20 \%$ of global CO$_2$ fossil fuel emissions, whilst lakes account for 9-24 $\%$  of CH$_4$ emissions, the second largest natural source after wetlands \citep{Saunois2020}. These rates of greenhouse gas emission are expected to rise further if the eutrophication of the Earth's lentic systems continues. With regards to weather, freezing and melting of the lake surface modifies the radiative and conductive properties and consequently affects the heat (latent, sensible) exchange and surface energy balance \citep{Huang2019,Peng2020,Franz2018}. Considering particular examples, over Lake Victoria convective activity is suppressed during the day and peaks at night, leading to intense, hazardous thunderstorms \citep{Thiery2015,Thiery_2017}; Lake Ladoga can generate low level clouds which can cause variability in the 2m temperature of up to 10 K \citep{Eerola2014}; the Laurentian Great Lakes can cause intense winter snow storms \citep{Vavrus2013} \citep{Notaro2013}. Moreover, as a result of the increased temperatures due to climate change, lakes become more numerous due to the melting of glaciers and permafrost. Additionally, the higher temperatures mean that previously permanent lake bodies become seasonal or intermittent. There is then evidently a huge potential return in the ability to accurately model the location, morphology and properties of lakes in weather and climate models. \newline

@@

\noindent It is a continuous enterprise to update the lake parametrization schemes and their input data fields to better represent small-scale surface processes. It is however challenging to accurately represent lakes in these parametrisations; the majority of lakes which are resolved at a 9km grid spacing have not had their morphology accurately measured, let alone monitored, whilst 28.9$\%$ of land and coastal cells are treated for sub-grid (\hl{i.e. covering half or less of a grid cell}) water. When introducing an updated lake representation it is difficult apriori to determine the additional value gained through doing so. There are two key factors here:
\begin{itemize}
	\item \hl{Are the updated fields closer to reality?}
	\item \hl{Do the updated fields increase the accuracy of the model predictions?} 
\end{itemize}
The first point is straightforward; we want our parametrisation fields to better represent reality. If the lake depth of some lake is updated from 10m to 100m we want to be sure that 100m is closer to the true depth of the lake. For the second point, even if the updated fields are accurate, are they informative in the sense that they enable us to make more accurate predictions? For instance, the main target of lake parametrization is to reproduce lake surface water temperatures (and therefore evaporation rates). If a lake parametrisation scheme is updated to better represent different types of inland waterbodies, the time variability of inland waterbodies and/or the lake morphology fields use more in situ measurements, does this additional information allow for more accurate predictions of the lake surface water temperatures? Is it therefore worthwhile to update the parametrisation in this way? Since the resulting updated fields are ultimately used operationally, it is essential to ensure the accuracy of the fields and prevent any potential degradation or instability of the model. This problem of quickly and automatically verifying the accuracy and information gain of updated lake parametrisations is the aim of this work. \newline

@@

\newpage
\section{Constructing VESPER}\label{sec:2}
\hl{In order to rapidly assess accuracy of new surface physiography fields and if their use in the model increase predictability a neural network regression model (further in the text referred as VESPER) that can learn the mapping between a set of features x and targets y is constructed. In this case the features are the atmospheric and surface model fields (such as 2 metre temperature from ERA5 reanalysis) and the surface physiographic fields (such as orography and vegetation cover used to produce ERA5 reanalysis), see Table 1 for the full list of variables used. The target is the satellite land surface temperature (skin temperature from MODIS Aqua Day MYD11A1 v006 collection).VESPER can then make predictions about the skin temperature given a set of input variables (i.e. atmospheric and surface model fields, and surface physiographic fields). In turn, these predictions can then be compared against observations (i.e. satellite skin temperature) and VESPER's accuracy evaluated. By varying the number/type and values of the input features x to VESPER and observing how the accuracy of its predictions change, some conclusions on if and how features can increase predictability of an actual atmospheric model. Moreover, by isolating geographic regions where the predictions get worse with new/updated surface physiographic fields, areas where these fields might be erroneous or not informative enough can be identified. Due to the inherent stochasticity of training a neural network regression model it is also possible for different models to settle in different local minimums i.e. the network variance.}

	\hl{To understand the significance of this, every VESPER configuration was trained four times, each time with a different seed. The size of this variance is much smaller than the variance between different VESPER configurations}

		In this section we will now describe the data used for the features x and targets y in the neural network regression model, how various data types are joined together, and the details of VESPER’s construction.

\subsection{Features and targets}

\hl{VESPER's input feature selection (see Table X) followed (i) permutation importance results for atmospheric and surface model fields - only fields with the highest importance were chosen; and (ii) expert choice for surface physiographic fields - as first attempt it was decided to test current methodology for lake related information, therefore fields that could be most affected by the presence of absence of water were selected, e.g. if lake had to be removed then some other surface had to appear (like bare ground, high or low vegetation, glacier or even ocean) and surface elevation had to change. Changes to the orographic fields will have important influences on temperature through e.g. wind, solar heating etc. Lake depth changes are similarly important, influencing how a lake freezes, thaws, mixes and its overall dynamical range. VESPER's target selection followed globally available criteria and the satellite land surface temperature is quite well observed globally and with high temporal pattern (daily or even several times a day depending on the location)}

\begin{table}[h]
	\begin{tabularx}{\textwidth}{lX}
		\hline
		Main surface physiographic fields
		(19 fields)          & Pressure: surface pressure (sp, Pa), mean sea level pressure (msl, Pa), 
		Wind: 10 metre U wind component (10u, m/s), 10 metre V wind component (10v, m/s), 
		Temperature: 2 metre temperature (2t, K), 2 metre dewpoint temperature (2d, K), skin temperature (skt, K), ice temperature layer 1 (the sea-ice temperature in layer 0-7 cm; istl1, K), ice temperature layer 2 (the sea-ice temperature in layer 7-28 cm; istl2, K), 
		Surface albedo: forecast albedo (fal, 0-1), 
		Snow: snow depth (sd, m of water equivalent)

		 \\
		Scenarie:               & At oprette en server med bestemte regler som tillader folk at spille sammen. More Text more text More Text \\
		\hline
	\end{tabularx}
\end{table}

\subsection{Data sources }
\hl{\textit{TK:This section is new}}

There are three main sources of data. The first is selection of surface physiographic fields from ERA5 \citep{Hersbach} and their updated versions \citep{Choulga2019,Boussetta2021} \citep{Mu2021} used as VESPER’s features.
The second is a selection of atmospheric and surface model fields from ERA5, also used as VESPER’s features. The third is day-time land surface temperature measurements from the Moderate Resolution Imaging Spectroradiometer (MODIS) onboard the Aqua satellite \citep{MODIS}, used as VESPER’s target variable.

\subsubsection{Surface physiographic fields}
\hl{\textit{TK:This entire section is new}}

\noindent Surface physiographic fields have gridded information of the Earth’s surface properties (e.g. land-use, vegetation type and distribution) and represent surface heterogeneity in the ECLand of ECMWF’s Integrated Forecasting System (IFS). They are used to compute surface turbulent fluxes (of heat, moisture and momentum) and skin temperature over different surfaces (vegetation, bare soil, snow, interception and water) and then to calculate an area-weighted average for the grid-box to couple with the atmosphere. To trigger all different parametrization schemes the ECMWF model uses a sets of physiographic fields, that do not depend on initial condition or forecast step. Most fields are constant; surface albedo is specified for 12 months to describe the seasonal cycle. Dependent on the origin, initial data comes at different resolutions and different projections, and is then first converted to a regular latitude-longitude grid (EPSG:4326) at ~1km at Equator resolution and secondly to a required grid and resolution. Surface physiographic fields used in this work consist of orographic, land, water, vegetation, soil, albedo fields and their difference between initial V15 and updated V20 field sets (and between expanded V15X and V20X field sets). See Tables 1 and 2 for the full list of surface physiographic fields and their input sources; for more details see IFS Documentation (2021).

As this work is focused on assessing quality of inland water information, main surface physiographic fields are lake cover (derived from land-sea mask) and lake mean depth (see Table 2). To generate V15 (and V15X) fractional lake cover the GlobCover2009 global map (Bontemps et al., 2011; Arino et al., 2012) is used. This map has a resolution of 300m, corresponds for the year 2009 and covers latitudes 85°N-60°S; corrections outside these latitudes for the polar regions are included separately. In the Arctic no land is assumed, in the Antarctic data from the high-resolution Radarsat Antarctic Mapping Project digital elevation model version 2 (RAMP2; Liu et al., 2015) is used. To generate V20 (and V20X) fractional lake cover more resent higher resolution datasets and updated methods have been used (Choulga et al., 2019). Main data source is the Joint Research Centre (JRC) the Global Surface Water Explorer (GSWE) dataset (Pekel et al., 2016). GSWE is a 30m resolution dataset from Landsat 5,7 and 8, providing information on the spatial and temporal variability of surface water on the Earth since March 1984; here only permanent water type was used for lake cover generation as it was more accurate inland water distribution on the annual basis (Choulga et al., 2019). Differences between V20 and V15 lake cover fields (see Figure 1) are consistent with the latest global and regional information: (i) increase of lake fraction in V20 compared to V15 over northern latitudes is due to permafrost melt leading to a new thermokarst lake emergence, and due to higher resolution input source and its better satellite image recognition methodologies; (ii) reduction of lake fraction in V20 compared to V15 can be explained with several reasons, like anthropogenic land use change (e.g. Aral Sea, which lies across the border between Uzbekistan and Kazakhstan, has been shrinking at an accelerated rate since the 1960s and started to stabilise in 2014 with an area of 7660km2, 9 times smaller than its size in 1960. GlobCover2009 describes the Aral Sea in 1998, when it was still “only” two times smaller than its 1960 extent, whereas GSWE provides a more up to date map.), use of only permanent water (e.g. Australia, where GlobCover2009 over-represents inland water, as most of these lakes are highly ephemeral, e.g. the endorheic Kati Thanda–Lake Eyre fills only a few times per century. The GSWE updates to this region therefore include only generally permanent water, removing all seasonal and rare ephemeral water.), and change in the ocean and inland water separation algorithm (e.g. north-east of Russia).
To generate V15 (and V15X) lake mean depth (see Figure 2a) the Global Lake DataBase version 1 (GLDBv1; Kourzeneva et al., 2012) is used. GLDBv1 has a resolution of ~1km and is based on ~13000 lakes with in situ lake depth information; outside this dataset all missing data grid-cells (i.e. over ocean and land) have 25 meter value; field aggregation to a coarser resolution is done by averaging. Overestimation of lake depth in summer season can result in strong cold biases and in winter season – lack of ice formation. To generate V20 (and V20X) lake mean depth (see Figure 2b) an updated version GLDBv3 (Choulga et al., 2014) is used. GLDBv3 has the same resolution of ~1km, but is based on an increased number of lakes with in situ lake depth information by ∼1500 (in addition has bathymetry information over all Finnish navigable lakes), it introduces distinction between freshwater and saline lakes (this information is currently not used by FLake), and suggests the method to assess the depth for lakes without in situ observations using geological and climate type information; field aggregation to a coarser resolution is done by computing the most occurring value. Verification of GLDBv1 and GLDBv3 lake depths against 353 Finnish lake measurements shows that GLDBv3 exhibits a 52 \% bias reduction in mean lake depth values compared to GLDBv1 (Choulga et al., 2019). For a further details on lake distribution and depth, the representation of lakes by ECMWF in general see Choulga et al. (2019) and Boussetta et al. (2021).

To expand V15 and V20 lake description (to V15X and V20X respectively, see Table 2) their salinity and time variability information was generated. Even though static permanent water fits better to describe inland water distribution on average all year round, some areas (in Tropics especially) would really benefit from having monthly varying information as they have a very strong seasonal cycle, when size, shape and depth of a lake changes over the course of the year, leading to a significant changes in modelling the lake temperature response. Similarly, saline lakes behave very differently to fresh water lakes since increased salt concentrations affect the density, specific heat capacity, thermal conductivity, and turbidity, as well as evaporation rates, ice formation and ultimately the surface temperature. These two properties of time variability and salinity are often related; it is common for saline lakes to fill and dry out over the course of the season, which naturally also affects the relative saline concentration of the lake itself. To create a monthly varying lake cover first 12 monthly fractional land-sea masks based on JRC Monthly Water History v1.3 maps for 2010-2020 were created. Since the annual lake maps were created taking into account a lot of additional sources the extra condition on the monthly maps that the monthly water is equal or greater than permanent water distribution from fractional land-sea mask is enforced. To create an inland salt lake cover GLDBv3 salt lake list was used. First, in order to identify separate lakes on ~1km resolution lake cover (here max lake distribution based on monthly varying lake cover is used to ease further application), small sub-grid lakes and large lake coasts are masked, i.e. grid-cells that have water fraction less than 0.25. Next, number of connected grid-cells in each lake (i.e. connected with sides only) is computed. Then only lakes that have 100 and more connected grid-cells are vectorised, as at ERA5 resolution of ∼31km the grid-cells are quite large and can include a mixture of freshwater and saline lakes. Finally, saline lake vectors are selected by filtering vectors which have no saline lake point from GLDBv3 located – in total 147 large salt lake vectors, which were further used to filter non-saline lakes at ~1km resolution lake cover, finally aggregated to ~31km resolution. In the future it is planned to revisit this field and extend the list to include additional data.

\begin{figure}
	\subfloat[\label{fiig:Modistime1}]{\includegraphics[width=0.48\textwidth]{example-image-a}} \\
	\caption{\hl{\textit{TK: This figure is a placeholder for Figure 1 in the word doc}}} 
	\label{fig:example_figure_a}
\end{figure}

\begin{figure}
	\subfloat[\label{fiig:Modistime1}]{\includegraphics[width=0.48\textwidth]{example-image-b}} \\
	\caption{\hl{\textit{TK: This figure is a placeholder for Figure 2 in the word doc}}} 
	\label{fig:example_figure_b}
\end{figure}

\begin{table}[h]
	\begin{tabularx}{\textwidth}{lX}
		\hline
		Main surface physiographic fields
		(19 fields)          & Pressure: surface pressure (sp, Pa), mean sea level pressure (msl, Pa), 
		Wind: 10 metre U wind component (10u, m/s), 10 metre V wind component (10v, m/s), 
		Temperature: 2 metre temperature (2t, K), 2 metre dewpoint temperature (2d, K), skin temperature (skt, K), ice temperature layer 1 (the sea-ice temperature in layer 0-7 cm; istl1, K), ice temperature layer 2 (the sea-ice temperature in layer 7-28 cm; istl2, K), 
		Surface albedo: forecast albedo (fal, 0-1), 
		Snow: snow depth (sd, m of water equivalent)

		\\
		Scenarie:               & At oprette en server med bestemte regler som tillader folk at spille sammen. More Text more text More Text \\
		\hline
	\end{tabularx}
\caption{\hl{\textit{TK: This table is a placeholder for Table 2 in the word doc}}}
\end{table}

\subsubsection{ERA5}
\hl{\textit{TK:This section has been updated}}

>>>>

\title{Deep learning for quality control of surface physiographic fields using satellite Earth observations}

@@

%\introduction  %% \introduction[modified heading if necessary]
\section{Introduction}
\noindent Globally, there are $\sim$ 117 million lakes - defined as inland water bodies without lateral movement of water - making up around 3.7$\%$ of the Earth's land surface \citep{Verpoorter2014}. Their distribution is highly anisotropic, with the majority of lakes located between $45-75^{\circ}$N in the Boreal and Arctic regions. Lakes are highly important from the perspective of both numerical weather prediction and climate modelling as part of the EC-Earth model. For the latter, lakes generally influence the global carbon cycle as both sinks and sources of greenhouse gases; the majority of lakes are net heterotrophic, over saturated with CO$_2$ as a result of in lake respiration and so emit carbon into the atmosphere \citep{Pace2005,Tranvik2009}.  Total CO$_2$ emission from lakes is estimated at $1.25 - 2.30$ Pg of CO2-equivalents annually \citep{DelSontro2018}, nearly $20 \%$ of global CO$_2$ fossil fuel emissions, whilst lakes account for 9-24 $\%$  of CH$_4$ emissions, the second largest natural source after wetlands \citep{Saunois2020}. These rates of greenhouse gas emission are expected to rise further if the eutrophication of the Earth's lentic systems continues. With regards to weather, freezing and melting of the lake surface modifies the radiative and conductive properties and consequently affects the heat (latent, sensible) exchange and surface energy balance \citep{Huang2019,Peng2020,Franz2018}. Considering particular examples, over Lake Victoria convective activity is suppressed during the day and peaks at night, leading to intense, hazardous thunderstorms \citep{Thiery2015,Thiery_2017}; Lake Ladoga can generate low level clouds which can cause variability in the 2m temperature of up to 10 K \citep{Eerola2014}; the Laurentian Great Lakes can cause intense winter snow storms \citep{Vavrus2013} \citep{Notaro2013}. Moreover, as a result of the increased temperatures due to climate change, lakes become more numerous due to the melting of glaciers and permafrost. Additionally, the higher temperatures mean that previously permanent lake bodies become seasonal or intermittent. There is then evidently a huge potential return in the ability to accurately model the location, morphology and properties of lakes in weather and climate models. \newline

@@

\noindent It is a continuous enterprise to update the lake parametrization schemes and their input data fields to better represent small-scale surface processes. It is however challenging to accurately represent lakes in these parametrisations; the majority of lakes which are resolved at a 9km grid spacing have not had their morphology accurately measured, let alone monitored, whilst 28.9$\%$ of land and coastal cells are treated for sub-grid (i.e. covering half or less of a grid cell) water. When introducing an updated lake representation it is difficult apriori to determine the additional value gained through doing so. There are two key factors here:
\begin{itemize}
	\item Are the updated fields closer to reality?
	\item Do the updated fields increase the accuracy of the model predictions?
\end{itemize}
The first point is straightforward; we want our parametrisation fields to better represent reality. If the lake depth of some lake is updated from 10m to 100m we want to be sure that 100m is closer to the true depth of the lake. For the second point, even if the updated fields are accurate, are they informative in the sense that they enable us to make more accurate predictions? For instance, the main target of lake parametrization is to reproduce lake surface water temperatures (and therefore evaporation rates). If a lake parametrisation scheme is updated to better represent different types of inland waterbodies, the time variability of inland waterbodies and/or the lake morphology fields use more in situ measurements, does this additional information allow for more accurate predictions of the lake surface water temperatures? Is it therefore worthwhile to update the parametrisation in this way? Since the resulting updated fields are ultimately used operationally, it is essential to ensure the accuracy of the fields and prevent any potential degradation or instability of the model. This problem of quickly and automatically verifying the accuracy and information gain of updated lake parametrisations is the aim of this work. \newline

@@

\newpage
\section{Constructing VESPER}\label{sec:2}
In order to rapidly assess accuracy of new surface physiography fields and if their use in the model increase predictability a neural network regression model (further in the text referred as VESPER) that can learn the mapping between a set of features x and targets y is constructed. In this case the features are the atmospheric and surface model fields (such as 2 metre temperature from ERA5 reanalysis) and the surface physiographic fields (such as orography and vegetation cover used to produce ERA5 reanalysis), see Table 1 for the full list of variables used. The target is the satellite land surface temperature (skin temperature from MODIS Aqua Day MYD11A1 v006 collection).VESPER can then make predictions about the skin temperature given a set of input variables (i.e. atmospheric and surface model fields, and surface physiographic fields). In turn, these predictions can then be compared against observations (i.e. satellite skin temperature) and VESPER's accuracy evaluated. By varying the number/type and values of the input features x to VESPER and observing how the accuracy of its predictions change, some conclusions on if and how features can increase predictability of an actual atmospheric model. Moreover, by isolating geographic regions where the predictions get worse with new/updated surface physiographic fields, areas where these fields might be erroneous or not informative enough can be identified. Due to the inherent stochasticity of training a neural network regression model it is also possible for different models to settle in different local minimums i.e. the network variance. \newline

\noindent \hl{To understand the significance of this, every VESPER configuration was trained four times, each time with a different seed. The size of this variance is much smaller than the variance between different VESPER configurations} \textit{\textbf{TK:TBD}} \newline

\noindent In this section we will now describe the data used for the features $x$ and targets $y$ in the neural network regression model, how various data types are joined together, and the details of VESPER’s construction.

\subsection{Features and targets}
VESPER's input feature selection (see Table \ref{table:definitions}) followed (i) permutation importance results for atmospheric and surface model fields - only fields with the highest importance were chosen; and (ii) expert choice for surface physiographic fields - as first attempt it was decided to test current methodology for lake related information, therefore fields that could be most affected by the presence of absence of water were selected, e.g. if lake had to be removed then some other surface had to appear (like bare ground, high or low vegetation, glacier or even ocean) and surface elevation had to change. Changes to the orographic fields will have important influences on temperature through e.g. wind, solar heating etc. Lake depth changes are similarly important, influencing how a lake freezes, thaws, mixes and its overall dynamical range. VESPER's target selection followed globally available criteria and the satellite land surface temperature is quite well observed globally and with high temporal pattern (daily or even several times a day depending on the location)

\begin{table}[h]
	\begin{tabularx}{\textwidth}{lX}
		\hline
		Atmospheric and surface model fields (11 fields)
		     & \textbf{Pressure}: surface pressure (\textit{sp, Pa}), mean sea level pressure (\textit{msl, Pa}), \newline 
		\textbf{Wind}: 10 metre U wind component (\textit{10u, m/s}), 10 metre V wind component (\textit{10v, m/s}), \newline 
		\textbf{Temperature}: 2 metre temperature (\textit{2t, K}), 2 metre dewpoint temperature (\textit{2d, K}), skin temperature (\textit{skt, K}), ice temperature layer 1 (the sea-ice temperature in layer 0-7 cm; \textit{istl1, K}), ice temperature layer 2 (the sea-ice temperature in layer 7-28 cm; \textit{istl2, K}), \newline 
		\textbf{Surface albedo:} forecast albedo (\textit{fal, 0-1}), \newline 
	\textbf{Snow:} snow depth (\textit{sd, m} of water equivalent)

		\\
		Main surface physiographic fields
		(19 fields)          & \textbf{Orographic fields:} standard deviation of filtered subgrid orography (\textit{sdfor, m}), standard deviation of orography (\textit{sdor, m}), anisotropy of sub-gridscale orography (\textit{isir}, -), angle of sub-gridscale orography (\textit{anor, radians}), slope of sub-gridscale orography (\textit{slor, -}), geopotential (the gravitational potential energy of a unit mass, at a particular location, relative to mean sea level; at the surface of the Earth, this parameter shows the variation in geopotential (height) of the surface, and is referred to as the orography; \textit{z}, $m^{2} s^{-2}$), \newline  
		\textbf{Land fields:} land-sea mask (the proportion of land, as opposed to ocean or inland waters (i.e. lakes, reservoirs, rivers, coastal waters), in a grid-cell; \textit{lsm, 0-1}), glacier mask (the proportion of a grid-cell covered by glacier; \textit{glm, 0-1}), \newline 
		\textbf{Water fields:} lake cover (the proportion of a grid-cell covered by inland water bodies; \textit{cl, 0-1}), lake total depth (the mean depth of inland water bodies; \textit{dl, m}), \newline 
		\textbf{Vegetation fields:} low vegetation cover (\textit{cvl, 0-1}), high vegetation cover (\textit{cvh, 0-1}), type of low vegetation (\textit{tvl}, -), type of high vegetation (\textit{tvh}, -), \newline 
		\textbf{Soil fields:} soil type (\textit{slt}, -), \newline 
		\textbf{Albedo fields:} UV visible albedo for direct radiation (\textit{aluvp}, 0-1), UV visible albedo for diffuse radiation (\textit{aluvd}, 0-1), near IR albedo for direct radiation (\textit{alnip}, 0-1), near IR albedo for diffuse radiation (\textit{alnid}, 0-1)

		\\
		Additional surface physiographic fields:               & Difference for all main surface physiographic fields between V15 and V20 field sets, \newline 
		Difference between V20 static lake cover and monthly varying lake cover (12 maps in total),\newline 
		Saline lake cover (the proportion of a grid-cell covered by saline inland water bodies; units: 0-1)
		 \\
		\hline
	\end{tabularx}
\caption{Input features used for training the NN model VESPER; atmospheric model fields (time varying) were kept the same in all simulations,  surface  physiographic  fields  (static)  were  updated  when  going  from  the  original  data  based  on GlobeCover2009/GLDBv1 (V15 field set) to GSWE/GLDBv3 (V20 field set); in brackets are variables description (where needed), short name (according to the GRIB parameter database) and units.
}
\label{table:definitions}
\end{table}

\subsection{Data sources }

There are three main sources of data. The first is selection of surface physiographic fields from ERA5 \citep{Hersbach} and their updated versions \citep{Choulga2019,Boussetta2021,Mu2021} used as VESPER’s features.
The second is a selection of atmospheric and surface model fields from ERA5, also used as VESPER’s features. The third is day-time land surface temperature measurements from the Moderate Resolution Imaging Spectroradiometer (MODIS) onboard the Aqua satellite \citep{MODIS}, used as VESPER’s target variable.

\subsubsection{Surface physiographic fields}
\noindent Surface physiographic fields have gridded information of the Earth’s surface properties (e.g. land-use, vegetation type and distribution) and represent surface heterogeneity in the ECLand of ECMWF’s Integrated Forecasting System (IFS). They are used to compute surface turbulent fluxes (of heat, moisture and momentum) and skin temperature over different surfaces (vegetation, bare soil, snow, interception and water) and then to calculate an area-weighted average for the grid-box to couple with the atmosphere. To trigger all different parametrization schemes the ECMWF model uses a sets of physiographic fields, that do not depend on initial condition or forecast step. Most fields are constant; surface albedo is specified for 12 months to describe the seasonal cycle. Dependent on the origin, initial data comes at different resolutions and different projections, and is then first converted to a regular latitude-longitude grid (EPSG:4326) at ~1km at Equator resolution and secondly to a required grid and resolution. Surface physiographic fields used in this work consist of orographic, land, water, vegetation, soil, albedo fields and their difference between initial V15 and updated V20 field sets (and between expanded V15X and V20X field sets). See Tables \ref{table:definitions} and \ref{tab:datasources} for the full list of surface physiographic fields and their input sources; for more details see IFS documentation \citep{IFSdocs}. As this work is focused on assessing quality of inland water information, main surface physiographic fields are lake cover (derived from land-sea mask) and lake mean depth (see Table \ref{tab:datasources}).  \newline 
\begin{table}[h]
	\begin{tabularx}{\textwidth}{lXX}
		\toprule 
		Field Category & V15 & V20 \\
		\hline
		Orographic & SRTM30 Shuttle Radar Topography Mission over 60°N-60°S; GLOBE: Global Land One-km Base Elevation Project data over 90-60°N; RAMP2: high-resolution Radarsat Antarctic Mapping Project Digital Elevation Model version 2 data \citep{Liu2015} over 60-90°S; BPRC: Byrd Polar Research Center over Greenland; IS 50V: Digital Map Database of Iceland over Iceland & As V15, with corrections to to erroneous shift
		\\
		\hline 
		Land   & \textbf{Source lsm:} GlobCover2009 \citep{GLOBCOVER,arino2012glcm} over 85°N-60°S; RAMP2: high-resolution Radarsat Antarctic Mapping Project Digital Elevation Model version 2 data \citep{Liu2015} over 60-90°S; no land assumed over 90-85°N \newline  
		\textbf{Source glm:} GLCC: Global Land Cover Characteristics version 2.0 over 90°N-90°S except Iceland; Icelandic Meteorological Office (IMO) glacier mask 2013 over Iceland
		 & \textbf{Source lsm:} GSWE: Global Surface Water Explorer \citep{GSWE}; glm \newline 
		 \textbf{Source glm:} Norwegian Institute glacier data over Svalbard; Icelandic Meteorological Office (IMO) glacier mask 2017 over Iceland; GIMP: Greenland Ice Mapping Project data \citep{Howat2014} over Greenland; CryoSat-2 satellite glacier data \citep{Slater2018} over Antarctica (+ manual gap filling);  GLIMS: Global Land Ice Measurements from Space data \citep{glims} over rest of the globe \\
		\hline
				Water   & 
				\textbf{Source cl:} \textit{lsm} (ocean is separated at actual resolution by seeding and removing all connected grid-cells, includes the Caspian Sea, the Azov Sea, The American Great Lakes) \newline 
				\textbf{Source dl:} The Caspian Sea bathymetry; Global Relief Model ETOPO1 \citep{amante2009egrm} over the Great Lakes, the Azov Sea; GLDB: Global Lake DataBase version 1 \citep{Kourzeneva2012} over rest of the globe; 25 meters assumed over missing data grid-cells
		& \textbf{Source cl:} \textit{lsm} (ocean is separated at ~1km resolution by upgraded flooding algorithm following \cite{Choulga2019} \newline 
		\textbf{Source dl:} GEBCO: General Bathymetric Charts of the Ocean \citep{Weatherall2015} over the Caspian Sea and the Azov Sea; Global Relief Model ETOPO1 \citep{amante2009egrm} over the Great Lakes; GLDB: Global Lake DataBase version 3 \citep{Choulga2014} over rest of the globe; indirect estimates based on geological origin of lakes \citep{Choulga2014} over missing data grid-cells \\
		\hline 
		Vegetation & GLCC: Global Land Cover Characteristics version 1.2. Note that vegetation type represent only dominant type over grid-cell & As V15 \\ 
		\hline
		 Soil & DSMW: FAO/UNESCO Digital Soil Map of the world \citep{FAO}. Note that soil type represent only dominant type over grid-cell & As V15 \\ 
		 \hline 
		 Albedo & MODIS 5-year climatology \citep{SCHAAF2002135}; RossThickLiSparseReciprocal BRDF model. note that Albedo values represent snow free surface albedo & As V15 \\ 
		\bottomrule
	\end{tabularx}
	\caption{List of input datasets for the surface physiographic fields for V15 and V20 field sets. V15X and V20X are identical to V15 and V20 respectively, but with the addition of saline lake cover, and monthly varying lake cover fields. \hl{TK: is glacier mask binary in V15 and 0-1 in V20?}}
	\label{tab:datasources}
\end{table}
\noindent To generate V15 (and V15X) fractional lake cover the GlobCover2009 global map  \citep{GLOBCOVER,arino2012glcm} is used. This map has a resolution of 300m, corresponds for the year 2009 and covers latitudes 85°N-60°S; corrections outside these latitudes for the polar regions are included separately. In the Arctic no land is assumed, in the Antarctic data from the high-resolution Radarsat Antarctic Mapping Project digital elevation model version 2 (RAMP2; Liu et al., 2015) is used. To generate V20 (and V20X) fractional lake cover more resent higher resolution datasets and updated methods have been used \citep{Choulga2019}. Main data source is the Joint Research Centre (JRC) the Global Surface Water Explorer (GSWE) dataset \citep{GSWE}. GSWE is a 30m resolution dataset from Landsat 5,7 and 8, providing information on the spatial and temporal variability of surface water on the Earth since March 1984; here only permanent water type was used for lake cover generation as it was more accurate inland water distribution on the annual basis \citep{Choulga2019}. Differences between V20 and V15 lake cover fields (see Figure 1) are consistent with the latest global and regional information: (i) increase of lake fraction in V20 compared to V15 over northern latitudes is due to permafrost melt leading to a new thermokarst lake emergence, and due to higher resolution input source and its better satellite image recognition methodologies; (ii) reduction of lake fraction in V20 compared to V15 can be explained with several reasons, like anthropogenic land use change (e.g. Aral Sea, which lies across the border between Uzbekistan and Kazakhstan, has been shrinking at an accelerated rate since the 1960s and started to stabilise in 2014 with an area of 7660km2, 9 times smaller than its size in 1960. GlobCover2009 describes the Aral Sea in 1998, when it was still “only” two times smaller than its 1960 extent, whereas GSWE provides a more up to date map.), use of only permanent water (e.g. Australia, where GlobCover2009 over-represents inland water, as most of these lakes are highly ephemeral, e.g. the endorheic Kati Thanda–Lake Eyre fills only a few times per century. The GSWE updates to this region therefore include only generally permanent water, removing all seasonal and rare ephemeral water.), and change in the ocean and inland water separation algorithm (e.g. north-east of Russia). \newline

\noindent To generate V15 (and V15X) lake mean depth (see Figure \ref{example_figure_a}) the Global Lake DataBase version 1 \citep[GLDBv1;][]{Kourzeneva2012} is used. GLDBv1 has a resolution of ~1km and is based on ~13000 lakes with in situ lake depth information; outside this dataset all missing data grid-cells (i.e. over ocean and land) have 25 meter value; field aggregation to a coarser resolution is done by averaging. Overestimation of lake depth in summer season can result in strong cold biases and in winter season – lack of ice formation. To generate V20 (and V20X) lake mean depth (see Figure \ref{example_figure_b}) an updated version GLDBv3 \citep{Choulga2014} is used. GLDBv3 has the same resolution of ~1km, but is based on an increased number of lakes with in situ lake depth information by ∼1500 (in addition has bathymetry information over all Finnish navigable lakes), it introduces distinction between freshwater and saline lakes (this information is currently not used by FLake), and suggests the method to assess the depth for lakes without in situ observations using geological and climate type information; field aggregation to a coarser resolution is done by computing the most occurring value. Verification of GLDBv1 and GLDBv3 lake depths against 353 Finnish lake measurements shows that GLDBv3 exhibits a 52 \% bias reduction in mean lake depth values compared to GLDBv1 \citep{Choulga2019}. For a further details on lake distribution and depth, the representation of lakes by ECMWF in general see \cite{Choulga2019}  and \cite{Boussetta2021}. \newline

To expand V15 and V20 lake description (to V15X and V20X respectively, see Table \ref{tab:datasources}) their salinity and time variability information was generated. Even though static permanent water fits better to describe inland water distribution on average all year round, some areas (in Tropics especially) would really benefit from having monthly varying information as they have a very strong seasonal cycle, when size, shape and depth of a lake changes over the course of the year, leading to a significant changes in modelling the lake temperature response. Similarly, saline lakes behave very differently to fresh water lakes since increased salt concentrations affect the density, specific heat capacity, thermal conductivity, and turbidity, as well as evaporation rates, ice formation and ultimately the surface temperature. These two properties of time variability and salinity are often related; it is common for saline lakes to fill and dry out over the course of the season, which naturally also affects the relative saline concentration of the lake itself. To create a monthly varying lake cover first 12 monthly fractional land-sea masks based on JRC Monthly Water History v1.3 maps for 2010-2020 were created. Since the annual lake maps were created taking into account a lot of additional sources the extra condition on the monthly maps that the monthly water is equal or greater than permanent water distribution from fractional land-sea mask is enforced. To create an inland salt lake cover GLDBv3 salt lake list was used. First, in order to identify separate lakes on ~1km resolution lake cover (here max lake distribution based on monthly varying lake cover is used to ease further application), small sub-grid lakes and large lake coasts are masked, i.e. grid-cells that have water fraction less than 0.25. Next, number of connected grid-cells in each lake (i.e. connected with sides only) is computed. Then only lakes that have 100 and more connected grid-cells are vectorised, as at ERA5 resolution of ∼31km the grid-cells are quite large and can include a mixture of freshwater and saline lakes. Finally, saline lake vectors are selected by filtering vectors which have no saline lake point from GLDBv3 located – in total 147 large salt lake vectors, which were further used to filter non-saline lakes at ~1km resolution lake cover, finally aggregated to ~31km resolution. In the future it is planned to revisit this field and extend the list to include additional data.
